\bf Adapting the Chumbley Score to match striae on Land Engraved Areas of bullets

### 2018-08-07 doc/paper.Rmd
fixed reference

---
title: "Adapting the Chumbley Score to match striae on Land Engraved Areas of bullets"
authors:
# - affiliation: Department of Statistics, Iowa State University 
- name: Ganesh Krishnan,
  degree: Master of Science (M.Sc.)
  # thank: The authors gratefully acknowledge funding from CSAFE through cooperative agreement \# 70NANB15H176 between NIST and Iowa State University. We would also like to thank the efforts of CSAFE and Roy J carver Microscopy lab in scanning the Hamby-44 set and providing the scans to us.
# - affiliation: Department of Statistics and CSAFE, Iowa State University 
- name: Prof. Heike Hofmann,
  degree: Doctor of Philosophy (Ph.D.)
affiliation: Department of Statistics, Iowa State University and CSAFE (Center for Statistics and Applications in Forensic Evidence)
funding: The authors gratefully acknowledge funding from CSAFE through cooperative agreement \# 70NANB15H176 between NIST and Iowa State University. 
thank: We would like to thank the efforts of CSAFE and Roy J Carver Microscopy lab in scanning the Hamby-44 set and providing the scans to us. We would also like to thank NIST for providing the Hamby-44 bullets to us for scanning.
presentedat:  This work was presented in the form of a poster in the AFTE 2018 49th Annual Training Seminar, held from June 3rd to June 8th, 2018 in Charleston, West Virginia
biblio-style: vancouver
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
    template: template.tex
  html_document: default
blinded: 0
keywords:
- forensic science
- toolmark
- cross-correlation
- Mann-Whitney U statistic
- land engraved areas (LEAs)
- algorithm
- signatures 
- same-source problem
bibliography: bibliography
abstract: The same-source problem remains a major challenge in forensic toolmark and firearm examination. Technological advances in surface imaging allow measurements of 3D surfaces at previously unforeseen resolutions and enable digitized imaging. Here, we investigate the applicability of the Chumbley scoring method \citep{hadler}, developed for screwdriver markings, for same-source identification of bullet striae. We provide methods to identify parameters that minimize the error rates for matching of LEAs using Hamby datasets 44 and 252 measured by NIST and CSAFE. We suggest a remedial algorithm to alleviate the problem of failed tests in the method, increasing both the power of the test and reducing error rates. Type II error rates of the proposed method improve by more than one third (Type I error of 0.05) and are on average about 0.22. This puts the proposed method on similar footing as other single feature matching approaches in the literature. 
---

@@

Same-source analyses are a major part of a Forensic Toolmark Examiner's job. In current practice, examiners  make these comparisons by visual inspection under a comparison microscope and come to one of the following four conclusions: identification, inconclusive, elimination or unsuitable for examination \citep{afte-toolmarks1998}. These conclusions are made on the basis of "unique surface contours" of the two toolmarks being in "sufficient agreement" \citep{afte-toolmarks1998}. AFTE describes the term "sufficient agreement" as the possibility of another tool producing the markings under comparison, as practically impossible \citep{afte-toolmarks1998}. Potential subject bias in the assessment as well as the lack of specified error rates are the main points of criticisms  first raised by the National Research Council in 2009 \citep{NAS:2009} and later emphasized further by the President's Council of Advisors on Science and Technology \citep{pcast2016}.

Technological advances, such as profilometers and confocal microscopy allow to measure 3D surfaces in a high-resolution digitized form. This technology has become more accessible over the last decade, and has made its way into topological images of ballistics evidence, such as bullet lands and breech faces \citep{DeKinder1, DeKinder2, Bachrach1, vorburger2016}.
Digitized images of 3D surfaces of form the data basis of statistical analysis of toolmarks. A statistical approach based on data  removes both subjectivity from the assessment and allows a quantification of error rates for both false positive and false negative identifications.

@@

Analysis of these digitized markings require the use of statistical methods which can quantify the scientific mechanism of comparing markings and serve as basis for an error rate calculation.
\citet{manytoolmarks2} define a relative distance metric and use it as similarity measure between two toolmarks. \citet{manytoolmarks1} extract many small segments in the markings of two toolmarks and compare similarity using a maximum Pearson correlation coefficient. 
The Chumbley scoring method, first introduced by \citet{chumbley}, uses a similar but more extensive framework based on a Mann-Whitney U test of the resulting correlation coefficients. This approach is non-deterministic, because segments are chosen randomly.  \citep{hadler} make the score deterministic for each pair of toolmarks by choosing segments for comparison systematically. This approach also ensures independence between segments of striae.

\begin{table}[ht]
\centering
\resizebox{\textwidth}{!}{\begin{tabular}{rlrrr}
  \hline
Research paper & Method & Data Source & False Positives & False Negatives\\
  \hline
 \textbf{\citet{manytoolmarks1}} & Maximum Pearson & Screwdrivers & & \\ & Correlation &  & - & - \\ \hline
 \textbf{\citet{chumbley}} & Randomized & Screwdrivers & & \\ (Same-Surface Same-Angle) & Chumbley Score & & 2.3\% & 8.9\%  \\ \hline
  \textbf{\citet{afte-chumbley}} & Randomized & Slip-joint  &  & \\ & Chumbley Score &  & - & - \\ \hline 
  \textbf{\citet{hadler}} & Deterministic & Screwdrivers & & \\ (Same-Surface Same-Angle) & Chumbley Score & & 0\% & 6\%  \\ \hline
  \textbf{\citet{manytoolmarks2}} & Similarity Measure & Screwdrivers & & \\ (Different Surfaces-same angle) & Relative Distance Metric & & 5.9\% &  9.4\% \\ (Same Surfaces-same angle)&  & & 0.22\% &  0\% \\
   \hline
\end{tabular}}
\caption{\label{tab:toolmarks-ER} Error Rates in same-source toolmark analysis reported in the literature. The top four reported papers use some variation of the Chumbley-score method.}
\end{table}

@@

After removing sub-class structure, 
the Chumbley score is calculated in two steps: an optimization step and a validation step.
In the optimization step, the two markings are aligned horizontally such that within a pre-defined window of length $w_o$ the correlation between $x(t_1)$ and $y(t_2)$ is maximized:
\[
\left(t_1^o, t_2^o\right) = \mathop{\arg \max}\limits_{1 \le t_1 \le T_1-w_o, 1 \le t_2 \le T_2-w_o} \text{cor} \left(x^{w_o} (t_1), y^{w_o}(t_2) \right)
\]
This results in an optimal vertical (in-phase) shift of $t_1^o - t_2^o$ for aligning the two markings. We will denote the relative optimal locations as $t_1^*$ and $t_2^*$, where $t_i^* = t_i^o/(T_i-w_o)$ for $i=1,2$, such that $t_1^*, t_2^* \in [0,1]$. Once (sub-)class characteristics are removed, the relative optimal locations should be distributed according to a uniform distribution in $[0,1]$.

In the validation step,  two sets of windows of size $w_v$ are chosen from both markings,  \autoref[see][]{fig:win-comparison}. In the first set, pairs of windows are extracted from the two markings using the optimal vertical shift as determined in the first step, whereas for the second set the windows are extracted using a different (out-of-phase) shift.

>>>>

---
title: "Adapting the Chumbley Score to match striae on Land Engraved Areas of bullets"
authors:
# - affiliation: Department of Statistics, Iowa State University 
- name: Ganesh Krishnan,
  degree: Master of Science (M.Sc.)
  # thank: The authors gratefully acknowledge funding from CSAFE through cooperative agreement \# 70NANB15H176 between NIST and Iowa State University. We would also like to thank the efforts of CSAFE and Roy J carver Microscopy lab in scanning the Hamby-44 set and providing the scans to us.
# - affiliation: Department of Statistics and CSAFE, Iowa State University 
- name: Prof. Heike Hofmann,
  degree: Doctor of Philosophy (Ph.D.)
affiliation: Department of Statistics, Iowa State University and CSAFE (Center for Statistics and Applications in Forensic Evidence)
funding: The authors gratefully acknowledge funding from CSAFE through cooperative agreement \# 70NANB15H176 between NIST and Iowa State University. 
thank: We would like to thank the efforts of CSAFE and Roy J Carver Microscopy lab in scanning the Hamby-44 set and providing the scans to us. We would also like to thank NIST for providing the Hamby-44 bullets to us for scanning.
presentedat:  This work was presented in the form of a poster in the AFTE 2018 49th Annual Training Seminar, held from June 3rd to June 8th, 2018 in Charleston, West Virginia
biblio-style: vancouver
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
    template: template.tex
  html_document: default
blinded: 0
keywords:
- forensic science
- toolmark
- cross-correlation
- Mann-Whitney U statistic
- land engraved areas (LEAs)
- algorithm
- signatures 
- same-source problem
bibliography: bibliography
abstract: The same-source problem remains a major challenge in forensic toolmark and firearm examination. Technological advances in surface imaging allow measurements of 3D surfaces at previously unforeseen resolutions and enable digitized imaging. Here, we investigate the applicability of the Chumbley scoring method \cite{hadler}, developed for screwdriver markings, for same-source identification of bullet striae. We provide methods to identify parameters that minimize the error rates for matching of LEAs using Hamby datasets 44 and 252 measured by NIST and CSAFE. We suggest a remedial algorithm to alleviate the problem of failed tests in the method, increasing both the power of the test and reducing error rates. Type II error rates of the proposed method improve by more than one third (Type I error of 0.05) and are on average about 0.22. This puts the proposed method on similar footing as other single feature matching approaches in the literature. 
---

@@

Same-source analyses are a major part of a Forensic Toolmark Examiner's job. In current practice, examiners  make these comparisons by visual inspection under a comparison microscope and come to one of the following four conclusions: identification, inconclusive, elimination or unsuitable for examination \cite{afte-toolmarks1998}. These conclusions are made on the basis of "unique surface contours" of the two toolmarks being in "sufficient agreement" \cite{afte-toolmarks1998}. AFTE describes the term "sufficient agreement" as the possibility of another tool producing the markings under comparison, as practically impossible \cite{afte-toolmarks1998}. Potential subject bias in the assessment as well as the lack of specified error rates are the main points of criticisms  first raised by the National Research Council in 2009 \cite{NAS:2009} and later emphasized further by the President's Council of Advisors on Science and Technology \cite{pcast2016}.

Technological advances, such as profilometers and confocal microscopy allow to measure 3D surfaces in a high-resolution digitized form. This technology has become more accessible over the last decade, and has made its way into topological images of ballistics evidence, such as bullet lands and breech faces \cite{DeKinder1, DeKinder2, Bachrach1, vorburger2016}.
Digitized images of 3D surfaces of form the data basis of statistical analysis of toolmarks. A statistical approach based on data  removes both subjectivity from the assessment and allows a quantification of error rates for both false positive and false negative identifications.

@@

Analysis of these digitized markings require the use of statistical methods which can quantify the scientific mechanism of comparing markings and serve as basis for an error rate calculation.
In \cite{manytoolmarks2} a relative distance metric is defined and used as similarity measure between two toolmarks. This approach is expanded in \citet{manytoolmarks1}: a set of small segments in the markings of two toolmarks are extracted and  similarity is compared using a maximum Pearson correlation coefficient. 
The Chumbley scoring method, first introduced in \cite{chumbley}, uses a similar but more extensive framework based on a Mann-Whitney U test of the resulting correlation coefficients. This approach is non-deterministic, because segments are chosen randomly.  In \cite{hadler}  the score is fixed to be deterministic for each pair of toolmarks by choosing segments for comparison systematically. This approach also ensures independence between segments of striae.

\begin{table}[ht]
\centering
\resizebox{\textwidth}{!}{\begin{tabular}{rlrrr}
  \hline
Research paper & Method & Data Source & False Positives & False Negatives\\
  \hline
 \textbf{Faden et al. (2007)} \cite{manytoolmarks1} & Maximum Pearson & Screwdrivers & & \\ & Correlation &  & - & - \\ \hline
 \textbf{Chumbley et al. (2010)} \cite{chumbley} & Randomized & Screwdrivers & & \\ (Same-Surface Same-Angle) & Chumbley Score & & 2.3\% & 8.9\%  \\ \hline
  \textbf{Grieve et al. (2014)}
 \cite{afte-chumbley} & Randomized & Slip-joint  &  & \\ & Chumbley Score &  & - & - \\ \hline 
  \textbf{Hadler \& Morris (2017)} \cite{hadler} & Deterministic & Screwdrivers & & \\ (Same-Surface Same-Angle) & Chumbley Score & & 0\% & 6\%  \\ \hline
  \textbf{Bachrach et al. (2010)} \cite{manytoolmarks2} & Similarity Measure & Screwdrivers & & \\ (Different Surfaces-same angle) & Relative Distance Metric & & 5.9\% &  9.4\% \\ (Same Surfaces-same angle)&  & & 0.22\% &  0\% \\
   \hline
\end{tabular}}
\caption{\label{tab:toolmarks-ER} Error Rates in same-source toolmark analysis reported in the literature. The top four reported papers use some variation of the Chumbley-score method.}
\end{table}

@@

After removing sub-class structure, 
the Chumbley score is calculated in two steps: an optimization step and a validation step.
In the optimization step, the two markings are aligned horizontally such that within a pre-defined window of length $w_o$ the correlation between $x(t_1)$ and $y(t_2)$ is maximized:
\[
\left(t_1^o, t_2^o\right) = \mathop{\arg \max}\limits_{1 \le t_1 \le T_1-w_o, 1 \le t_2 \le T_2-w_o} \text{cor} \left(x^{w_o} (t_1), y^{w_o}(t_2) \right)
\]
This results in an optimal vertical (in-phase) shift of $t_1^o - t_2^o$ for aligning the two markings. We will denote the relative optimal locations as $t_1^*$ and $t_2^*$, where $t_k^* = t_k^o/(T_k-w_o)$ for $k=1,2$, such that $t_1^*, t_2^* \in [0,1]$. Once (sub-)class characteristics are removed, the relative optimal locations should be distributed according to a uniform distribution in $[0,1]$.

In the validation step,  two sets of windows of size $w_v$ are chosen from both markings, see \autoref{fig:win-comparison}. In the first set, pairs of windows are extracted from the two markings using the optimal vertical shift as determined in the first step, whereas for the second set the windows are extracted using a different (out-of-phase) shift.

### 2018-08-07 doc/paper.tex
fixed reference

\bigskip
\begin{abstract}
The same-source problem remains a major challenge in forensic toolmark
and firearm examination. Technological advances in surface imaging allow
measurements of 3D surfaces at previously unforeseen resolutions and
enable digitized imaging. Here, we investigate the applicability of the
Chumbley scoring method \citep{hadler}, developed for screwdriver
markings, for same-source identification of bullet striae. We provide
methods to identify parameters that minimize the error rates for
matching of LEAs using Hamby datasets 44 and 252 measured by NIST and
CSAFE. We suggest a remedial algorithm to alleviate the problem of
failed tests in the method, increasing both the power of the test and
reducing error rates. Type II error rates of the proposed method improve
by more than one third (Type I error of 0.05) and are on average about
0.22. This puts the proposed method on similar footing as other single
feature matching approaches in the literature.
\end{abstract}

@@

Same-source analyses are a major part of a Forensic Toolmark Examiner's
job. In current practice, examiners make these comparisons by visual
inspection under a comparison microscope and come to one of the
following four conclusions: identification, inconclusive, elimination or
unsuitable for examination \citep{afte-toolmarks1998}. These conclusions
are made on the basis of ``unique surface contours'' of the two
toolmarks being in ``sufficient agreement'' \citep{afte-toolmarks1998}.
AFTE describes the term ``sufficient agreement'' as the possibility of
another tool producing the markings under comparison, as practically
impossible \citep{afte-toolmarks1998}. Potential subject bias in the
assessment as well as the lack of specified error rates are the main
points of criticisms first raised by the National Research Council in
2009 \citep{NAS:2009} and later emphasized further by the President's
Council of Advisors on Science and Technology \citep{pcast2016}.

Technological advances, such as profilometers and confocal microscopy
allow to measure 3D surfaces in a high-resolution digitized form. This
technology has become more accessible over the last decade, and has made
its way into topological images of ballistics evidence, such as bullet
lands and breech faces
\citep{DeKinder1, DeKinder2, Bachrach1, vorburger2016}. Digitized images
of 3D surfaces of form the data basis of statistical analysis of
toolmarks. A statistical approach based on data removes both
subjectivity from the assessment and allows a quantification of error
rates for both false positive and false negative identifications.

@@

After removing sub-class structure, the Chumbley score is calculated in
two steps: an optimization step and a validation step. In the
optimization step, the two markings are aligned horizontally such that
within a pre-defined window of length \(w_o\) the correlation between
\(x(t_1)\) and \(y(t_2)\) is maximized: \[
\left(t_1^o, t_2^o\right) = \mathop{\arg \max}\limits_{1 \le t_1 \le T_1-w_o, 1 \le t_2 \le T_2-w_o} \text{cor} \left(x^{w_o} (t_1), y^{w_o}(t_2) \right)
\] This results in an optimal vertical (in-phase) shift of
\(t_1^o - t_2^o\) for aligning the two markings. We will denote the
relative optimal locations as \(t_1^*\) and \(t_2^*\), where
\(t_i^* = t_i^o/(T_i-w_o)\) for \(i=1,2\), such that
\(t_1^*, t_2^* \in [0,1]\). Once (sub-)class characteristics are
removed, the relative optimal locations should be distributed according
to a uniform distribution in \([0,1]\).

In the validation step, two sets of windows of size \(w_v\) are chosen
from both markings, \autoref[see][]{fig:win-comparison}. In the first
set, pairs of windows are extracted from the two markings using the
optimal vertical shift as determined in the first step, whereas for the
second set the windows are extracted using a different (out-of-phase)
shift.

>>>>

\bigskip
\begin{abstract}
The same-source problem remains a major challenge in forensic toolmark
and firearm examination. Technological advances in surface imaging allow
measurements of 3D surfaces at previously unforeseen resolutions and
enable digitized imaging. Here, we investigate the applicability of the
Chumbley scoring method \cite{hadler}, developed for screwdriver
markings, for same-source identification of bullet striae. We provide
methods to identify parameters that minimize the error rates for
matching of LEAs using Hamby datasets 44 and 252 measured by NIST and
CSAFE. We suggest a remedial algorithm to alleviate the problem of
failed tests in the method, increasing both the power of the test and
reducing error rates. Type II error rates of the proposed method improve
by more than one third (Type I error of 0.05) and are on average about
0.22. This puts the proposed method on similar footing as other single
feature matching approaches in the literature.
\end{abstract}

@@

Same-source analyses are a major part of a Forensic Toolmark Examiner's
job. In current practice, examiners make these comparisons by visual
inspection under a comparison microscope and come to one of the
following four conclusions: identification, inconclusive, elimination or
unsuitable for examination \cite{afte-toolmarks1998}. These conclusions
are made on the basis of ``unique surface contours'' of the two
toolmarks being in ``sufficient agreement'' \cite{afte-toolmarks1998}.
AFTE describes the term ``sufficient agreement'' as the possibility of
another tool producing the markings under comparison, as practically
impossible \cite{afte-toolmarks1998}. Potential subject bias in the
assessment as well as the lack of specified error rates are the main
points of criticisms first raised by the National Research Council in
2009 \cite{NAS:2009} and later emphasized further by the President's
Council of Advisors on Science and Technology \cite{pcast2016}.

Technological advances, such as profilometers and confocal microscopy
allow to measure 3D surfaces in a high-resolution digitized form. This
technology has become more accessible over the last decade, and has made
its way into topological images of ballistics evidence, such as bullet
lands and breech faces
\cite{DeKinder1, DeKinder2, Bachrach1, vorburger2016}. Digitized images
of 3D surfaces of form the data basis of statistical analysis of
toolmarks. A statistical approach based on data removes both
subjectivity from the assessment and allows a quantification of error
rates for both false positive and false negative identifications.

@@

After removing sub-class structure, the Chumbley score is calculated in
two steps: an optimization step and a validation step. In the
optimization step, the two markings are aligned horizontally such that
within a pre-defined window of length \(w_o\) the correlation between
\(x(t_1)\) and \(y(t_2)\) is maximized: \[
\left(t_1^o, t_2^o\right) = \mathop{\arg \max}\limits_{1 \le t_1 \le T_1-w_o, 1 \le t_2 \le T_2-w_o} \text{cor} \left(x^{w_o} (t_1), y^{w_o}(t_2) \right)
\] This results in an optimal vertical (in-phase) shift of
\(t_1^o - t_2^o\) for aligning the two markings. We will denote the
relative optimal locations as \(t_1^*\) and \(t_2^*\), where
\(t_k^* = t_k^o/(T_k-w_o)\) for \(k=1,2\), such that
\(t_1^*, t_2^* \in [0,1]\). Once (sub-)class characteristics are
removed, the relative optimal locations should be distributed according
to a uniform distribution in \([0,1]\).

In the validation step, two sets of windows of size \(w_v\) are chosen
from both markings, see \autoref{fig:win-comparison}. In the first set,
pairs of windows are extracted from the two markings using the optimal
vertical shift as determined in the first step, whereas for the second
set the windows are extracted using a different (out-of-phase) shift.
